---
title: "Flooding and High Freq RB notes"
author: "Rachel Bash"
date: "11/10/2019"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

pacman::p_load(tidyverse, dataRetrieval, dygraphs, maps, sf, xts)

getwd()
highfreqsiteinfo <- 
  read.csv("./Data/Raw/highfreqsiteinfo.csv", as.is=TRUE)

highfreqsite2019 <- highfreqsiteinfo %>%
  filter(end_date > "2019-03-31") #4 sites have high freq N data in 2019

#read in the data for 4 sites
highfreqsites.DN <- readNWISuv(site = c("06808500", "06817000", "06892350", "06934500"), 
                               parameterCd = c("00060", "99133"), 
                               # Discharge in cfs & Nitrate in mg/l NO3-N
                               startDate = "2019-01-01",
                               endDate = "2019-11-01") %>%
                               renameNWISColumns() %>%
                               rename(Nitrate_mgl = 6)
highfreqsites2019 <- write.csv(highfreqsites.DN, "./Data/Raw/highfreqsite2019.csv")

#individual sites
Hermann <- highfreqsites.DN %>%
           filter(site_no=="06934500")
Desoto <- highfreqsites.DN %>%
          filter(site_no=="06892350")
Clarinda <- highfreqsites.DN %>%
            filter(site_no=="06817000")
Randolph <- highfreqsites.DN %>%
            filter(site_no=="06808500")

library(cowplot)
```

<<<<<<< HEAD
All directions from Kateri in <>
< Setup the global options for the R chunks in your document >

<Note: set up autoreferencing for figures and tables in your document.>

<Paragraph detailing the rationale for your analysis. What is the significant application and/or interest in this topic? Connect to environmental topic(s)/challenge(s).>

<Paragraph detailing your research question(s) and goals. What do you want to find out? Include a sentence (or a few) on the dataset you are using to answer this question - just enough to give your reader an idea of where you are going with the analysis.>

<Information on how the dataset for this analysis were collected, the data contained in the dataset, and any important pieces of information that are relevant to your analyses. This section should contain much of same information as the README file for the dataset but formatted in a way that is more narrative.>

<Add a table that summarizes your data structure. This table can be made in markdown text or inserted as a `kable` function in an R chunk. If the latter, do not include the code used to generate your table.>

<C will do data table for water quality and daily values, R will do for high freq>

<Include R chunks for 5+ lines of summary code (display code and output), 3+ exploratory graphs (display graphs only), and any wrangling you do to your dataset(s).> 

<Include text sections to accompany these R chunks to explain the reasoning behind your workflow, and the rationale for your approach.>

<Include R chunks for 3+ statistical tests (display code and output) and 3+ final visualization graphs (display graphs only).>

<Include text sections to accompany these R chunks to explain the reasoning behind your workflow, rationale for your approach, and the justification of meeting or failing to meet assumptions of tests.>

=======

```{r}
highfreqsites1718 <- readNWISuv(site = c("06808500", "06817000", "06892350", 
                                         "06934500", "06892513", "06902000"), 
                               parameterCd = c("00060", "99133"), 
                               # Discharge in cfs & Nitrate in mg/l NO3-N
                               startDate = "2017-01-01",
                               endDate = "2019-01-01") %>%
                               renameNWISColumns() %>%
                               rename(Nitrate_mgl = 6)

ggplot(highfreqsites1718, aes(x=dateTime, y=Nitrate_mgl, color = site_no)) +
  geom_line()

Hermann <- highfreqsites1718 %>%
           filter(site_no=="06934500")
Desoto <- highfreqsites1718 %>%
          filter(site_no=="06892350")
Clarinda <- highfreqsites1718 %>%
            filter(site_no=="06817000")
Randolph <- highfreqsites1718 %>%
            filter(site_no=="06808500")
Sumner <- highfreqsites1718 %>%
            filter(site_no=="06902000")
Shawnee <- highfreqsites1718 %>%
            filter(site_no == "6892513")
```


```{r}
dygraph(
  cbind(
    Flow = xts(Hermann$Flow_Inst, order.by = Hermann$dateTime), 
    Nitrate = xts(Hermann$Nitrate_mgl, order.by = Hermann$dateTime)
  )
) %>% 
  dySeries("Nitrate", axis = "y2") %>%
  dyRangeSelector() #Oct

dygraph(
  cbind(
    Flow = xts(Desoto$Flow_Inst, order.by = Desoto$dateTime), 
    Nitrate = xts(Desoto$Nitrate_mgl, order.by = Desoto$dateTime)
  )
) %>% 
  dySeries("Nitrate", axis = "y2") %>%
  dyRangeSelector() #Dec

dygraph(
  cbind(
    Flow = xts(Clarinda$Flow_Inst, order.by = Clarinda$dateTime), 
    Nitrate = xts(Clarinda$Nitrate_mgl, order.by = Clarinda$dateTime)
  )
) %>% 
  dySeries("Nitrate", axis = "y2") %>%
  dyRangeSelector() #mid Oct

dygraph(
  cbind(
    Flow = xts(Randolph$Flow_Inst, order.by = Randolph$dateTime), 
    Nitrate = xts(Randolph$Nitrate_mgl, order.by = Randolph$dateTime)
  )
) %>% 
  dySeries("Nitrate", axis = "y2") %>%
  dyRangeSelector() #mid Oct

dygraph(
  cbind(
    Flow = xts(Sumner$Flow_Inst, order.by = Sumner$dateTime), 
    Nitrate = xts(Sumner$Nitrate_mgl, order.by = Sumner$dateTime)
  )
) %>% 
  dySeries("Nitrate", axis = "y2") %>%
  dyRangeSelector() #beg sept

dygraph(
  cbind(
    Flow = xts(Shawnee$Flow_Inst, order.by = Shawnee$dateTime), 
    Nitrate = xts(Shawnee$Nitrate_mgl, order.by = Shawnee$dateTime)
  )
) %>% 
  dySeries("Nitrate", axis = "y2") %>%
  dyRangeSelector()
```
>>>>>>> f971452bbc172a905cf2296de78d7c94481e471d

## Sites


There were 7 sites in our region of interest that had high freq N data, and only 4 sites had high freq N data during the floods of 2019. The sites looked at in depth are:

    - West Nishnabotna River in Randolph, IA
    - Nodaway River at Clarinda, IA
    - Kansas River in Desoto, KS
    - Missouri River at Hermann, MO

The Missouri River is the biggest river, with an average of 214693 cfs discharge rate during the year 2019, and the Nodaway River is the smallest river, with an average of 1185 cfs discharge rate for 2019.

## C-Q plots and hysteresis plots

In March of 2019, a [bomb cyclone](https://www.kansascity.com/news/state/missouri/article228237519.html) hit the midwest. Our initial research question, what effect did the March 2019 storm have on water quality, attempted to look into the behavior of nitrogen in the discharge of the rivers. Unfortunately, instantaneous Nitrogen values stopped recording during the peak of the storm events in March, so it was hard to create hysteresis plots that exhibited the type of storm and its effects on nitrogen concentration. 

Even though Nitrogen concentrations were not recorded in March, they were recorded in other times of the year. 2019 was a wet year and many large storm events occurred. 

```{r randolphs, echo=FALSE}
#Randolph storm
RandolphStorm <- Randolph %>%
  filter(dateTime > "2019-06-24" & dateTime < "2019-07-08") 


RandolphStorm.plot <- ggplot(RandolphStorm, aes(x = Flow_Inst, y = Nitrate_mgl, color = dateTime)) +
  geom_point() +
  labs(x="Discharge (cfs)", y= "Nitrogen mg/l)", color="Date", 
       title="West Nishnabotna River in Randolph, IA")
print(RandolphStorm.plot)
#semi-clockwise motion, negative slope, so diluting
```

```{r desotos, fig.cap="\\label{fig:desotos} Hysteresis plots for two storm events in the Missouri River Basin. The storm event on the West Nishnabotna River exhibits an oddly-shaped plot that has a negative slope, indicating it is a diluting storm. The Kansas River experienced a storm in late February that has a counter-clockwise motion and a positive slope, indicating a flushing storm. These two plots illustrate that two rivers near each other can have very different behaviors."}
#desoto storm

DesotoStorm <- Desoto %>%
  filter(dateTime > "2019-02-22" & dateTime < "2019-02-28") 


DesotoStorm.plot <- ggplot(DesotoStorm, aes(x = Flow_Inst, y = Nitrate_mgl, color = dateTime)) +
  geom_point() +
  labs(x="Discharge (cfs)", y= "Nitrogen mg/l)", color="Date", 
       title="Kansas River in Desoto, KS")


plot_grid(RandolphStorm.plot, DesotoStorm.plot, ncol=1)
#very loose counter clockwise motion - loose positive slope for flushing storm
```



## Baseflow

Baseflow and quickflow were separated for the 4 rivers with high frequency discharge data during the year 2019 and it was clear that all the rivers have a majority of their water coming from baseflow. 

River            | % Discharge coming from baseflow
---------------- | ---------------------------------
Missouri River   | 98.9%
Nodaway River    | 82.5%
Kansas River     | 96.2%
West Nishnabotna | 94.9%


The percent of discharge coming from baseflow correlates with the general size of the river. If the river is large, more of the discharge will come from baseflow. The West Nishnabotna River in Randolph, Iowa was further examined to see whether the baseflow percentage changes depending on the storm event. The data was reduced and only discharge data from the month of March in 2019 was examined. Total baseflow export was 91.1%, which is lower, but not that much lower than the baseflow separation for the whole 2019 year.




## 2018 October flood
Because I was disappointed in my lack of results, I chose to also look at the 2018 flood in October. This wasn't as big as the one in March.

There were clearer hysteresis plots that were able to be produced. Clarinda, Randolph, and Shawnee had counterclockwise, diluting storms with negative slope. Meaning the quickflow had low concentration of Nitrate. 


#Questions
- Land use: how do use?
- Can I look at a storm event in a river and assume that nitrogen will behave the same way every time during that storm event?
- Look at definition of drought to see if there have been more frequent occurances of drought over time?
7Q10 look at 10 years of data. what is average min 7 day flow. Avg low flowPath. 
- if discharge is below 
- look at daily values for C-Q plots to see whether systems are flushing versus diluting

```{r}
library(sf)
getwd()

CDL2018<- foreign::read.dbf("C:/Users/19524/Downloads/polygonclip_20191115123758_91314621/CDL_2018_clip.tif.vat.dbf", as.is=TRUE)




```


# C-Q with daily values

```{r}
bestsites.DNP <- read.csv("./Data/Raw/bestsites.DNP.csv")
unique(bestsites.DNP$parm_cd)

bestsites.DNP.skinny <- bestsites.DNP %>%
  dplyr::select(Site = site_no,
         Date = Date,
         Parameter = parm_cd,
         Value = result_va,
         Discharge = Discharge) %>%
  group_by(Date, Parameter, Site) %>%
  summarize(Value = mean(Value),
            Discharge = mean(Discharge)) %>%
  spread(key = Parameter, value = Value) %>% 
  rename(Discharge2 = '60', total.nitrogen = '600', 
         total.phosphorus = '665') %>%
  mutate(Year = year(Date))

ggplot(bestsites.DNP.skinny, aes(x=Discharge2, y= total.nitrogen)) +
  geom_point()
```

